### 2020-06-17 report/sections/server/node.tex
Report: update graphic placement
Report: complete chapter 4 sans connection
Report: review chapter 4

\chapter{\fancyname{NodeMPST}: Back-End Session Type Web Development}
\label{chap:node}

In this chapter we present \fancyname{NodeMPST},
our session type API generation strategy for 
server-side endpoints implemented on the Node.js runtime \cite{node}.
We continue to use the running example of the \tprotocol{Adder}
protocol, and refer to the FSM of the \trole{Svr} endpoint
(\cref{fig:addersvrfsm}) throughout this chapter.

\begin{figure}[!b]
\centering
\includegraphics[width=0.5\textwidth]{AdderSvrFSM}
\captionof{figure}{\trole{Svr} Endpoint FSM
in \tprotocol{Adder} protocol}
\label{fig:addersvrfsm}
\end{figure}

@@

>>>>

\chapter{\nodecodegen: Back-End Session Type Web Development}
\label{chap:node}

In this chapter we present \nodecodegen,
our session type API generation strategy for 
server-side endpoints implemented on the Node.js runtime 
\cite{nodejs}.
We continue to use the running example of the \tprotocol{Adder}
protocol (\cref{lst:adder}),
and refer to the FSM of the \trole{Svr} endpoint
(\cref{fig:addersvrfsm}) throughout this chapter.

We discuss the challenges of implementing
session types on the Node.js runtime 
(\cref{section:nodechallenges})
and motivate our approach (\cref{section:nodeapproach}).
We explain our design choices for encoding
the EFSM for server-side endpoints (\cref{section:nodeefsm}),
and detail the session runtime that is designed
to execute the EFSM on Node.js (\cref{section:noderuntime}).
We conclude by analysing some alternative designs 
(\cref{section:nodealt}) and evaluating the limitations 
of our approach (\cref{section:nodelimitations}).

\begin{figure}[!b]
\centering
\includegraphics[width=0.6\textwidth]{AdderSvrFSM}
\captionof{figure}{\trole{Svr} Endpoint FSM
in \tprotocol{Adder} protocol}
\label{fig:addersvrfsm}
\end{figure}

@@

\section{Summary}
In this chapter, we have presented
our session type API generation strategy
which targets the Node.js runtime for server-side
endpoints.

We motivated our approach of generating
handler-style APIs that act as seams in the session runtime
which executes the EFSM.
By doing so, the mechanism for sending and receiving
messages are not exposed to the developer,
which makes channel reuse impossible by construction.

We discussed our session runtime implementation in great
depth and iterated upon the design to arrive at 
a minimal runtime implementation that executes the EFSM
using a message-passing abstraction internally as well.
In particular, we motivated how to handle message receives
in a way that preserves the message ordering specified
in the communication protocol.

Developers that implement their Node.js endpoint application
using the generated APIs enjoy protocol conformance
by construction: 
the TypeScript type system will prevent type mismatches
in the handlers defined by the developer
(e.g. sending a \lstonelinejs{string} instead of 
a \lstonelinejs{number}, or using a message label
that is not defined on the EFSM state),
and the handler-style APIs themselves conceal channel
resources to prevent channel reuse.

### 2020-06-17 report/sections/server/node/approach.tex
Report: review chapter 4

We motivate our approach from \cite{PureScript2019, HuJava}
to generate \textbf{handler-style APIs} to be implemented
by developers.
This aligns with idiomatic TypeScript practices of defining
\textit{callbacks} in the application logic.
We type the parameters and return values of the handlers to
reflect the message types specified in the protocol.
By strictly specifying handlers for send and receive actions,
we do not expose send and receive APIs to the developer,
making it impossible for the developer to reuse channels.

@@

When executing \fancyname{SessionTS} to generate code for the
\trole{Svr} endpoint specifying the \texttt{node} target,
the developer obtains the following files:

>>>>

We motivate our approach from \cite{PureScript2019, HuJava}
to generate \textbf{handler-style APIs} to be implemented
by developers.
This aligns with idiomatic TypeScript practices of defining
\textit{callbacks} in the application logic.
We type the parameters and return values of the handlers to
reflect the message types specified in the protocol.
By strictly specifying handlers for send and receive actions,
we do not expose send and receive APIs to the developer,
contrary to \cite{HuJava},
making it impossible for the developer to reuse channels,
hence saving the need for linearity checks.

@@

When executing \nodecodegen to generate code for the
\trole{Svr} endpoint specifying the \texttt{node} target,
the developer obtains the following files:

### 2020-06-17 report/sections/server/node/challenges.tex
Report: review chapter 4

Our goal with \fancyname{NodeMPST} is to
generate session type implementations that
provide as much \textbf{static} communication safety guarantees
as possible,
in a manner suitable for both the TypeScript language 
and the single-threaded event loop model in Node.js
As outlined in \cref{section:typescript}, TypeScript's type system
is neither affine nor linear, so we need to work around the 
language's limitations to enforce linear channel usage.

>>>>

Our goal with \nodecodegen is to
generate session type implementations that
provide as much \textbf{static} communication safety guarantees
as possible,
in a manner suitable for both the TypeScript language 
and the single-threaded event loop model in Node.js
TypeScript's type system
is neither affine nor linear, so we need to work around the 
language's limitations to enforce linear channel usage.

### 2020-06-17 report/sections/server/node/efsm.tex
Report: progress on chapter 4 receive edge case
Report: complete chapter  4 runtime
Report: minor changes to chapter 4
Report: review chapter 4

We show the structure of the generated EFSM.ts file in
\cref{lst:nodeefsmfile}.
Note that the formal definition of the EFSM in 
\cref{section:scribbleefsm}
contains more than just states and the state transition function,
so we encode the additional information as well.
Each type of information is grouped into their own
\textit{namespace}, and are collectively exported in
the EFSM \textit{module} for the developer to use.

\begin{figure}
\begin{lstlisting}[language=javascript,tabsize=2,title=EFSM.ts]
// (*@\cref{subsection:nodeefsmroleslabelsmsg}@*)
export namespace Roles {...};
export namespace Labels {...};
export namespace Message {...};

// (*@\cref{subsection:nodeefsmhandlers}@*)
export namespace Handler {...};

// (*@\cref{subsection:nodeefsmimplementation}@*)
abstract class ISend {...};
abstract class IReceive {...};
abstract class ITerminal {...};
export namespace Implementation {...};

export type EfsmTransitionHandler =
	(implementation: Implementation.Type) => void;
export type MessageHandler = (message: any) => void;
\end{lstlisting}
\captionof{lstlisting}{Structure of Generated EFSM Encoding 
for Server Endpoint}
\label{lst:nodeefsmfile}
\end{figure}

@@

\subparagraph{Messages}
The handler APIs that we generate for developers
need to refer to the label identifier and payload type: 
we refer to this as the message structure, and encode this as a
Message Type. 
Each message is expressed as an interface with
properties for the label and payload.
These interfaces are grouped based on the EFSM state
they belong using \textit{union types}.
We illustrate this in \cref{lst:addersvrmsg}.
By expressing the payload type as a \textit{tuple}\footnote{
In TypeScript, a tuple is an array with fixed size
and known types for elements at each position.
},
we easily generalise our type definition to polyadic payloads.

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript, tabsize=2]
// Inside the Message namespace...
export interface S54THANKS {
	label: Labels.S54.THANKS,
	payload: [string],
};
export interface S54TERMINATE {
	label: Labels.S54.TERMINATE,
	payload: [],
};

@@

We collect the APIs that the developer needs to implement
under the \texttt{Handler} namespace. 
As a design choice, we \textit{do not} generate handlers for
terminal states, because the semantics of inactivity mean
there is nothing to handle.
We introduce the generated handlers for sending and receiving states.
These are non-terminal states that will involve the encoding of its
successor state. The reader will notice that, in the listings below,
the successor state is stated to be under the
\texttt{Implementation} namespace: we explain in 
\cref{subsection:nodeefsmimplementation}, but for now,
it is sufficient to acknowledge that those refer to the encoding
of the successor state.

\subparagraph{Send}
We model selections using a union type to
encapsulate the possible send actions, as shown in 
\cref{lst:addersvrsendhandler}.
Each send action is encoded as a tuple of
the label, the payload, and the successor state encoding.
We see some benefits from defining Message Types as interfaces:
TypeScript supports \textit{index type queries} to extract
named property types, so
\lstonelinejs{Message.S54THANKS['payload']} 
would resolve to \lstonelinejs{[string]},
based on the interface definition \cref{lst:addersvrmsg}.

@@

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript,tabsize=2]
// Inside the Handler namespace...
export type S51 = {
	[Labels.S51.ADD]: (...payload: Message.S51ADD['payload']) =>
		Implementation.S53,
	[Labels.S51.QUIT]: (...payload: Message.S51QUIT['payload']) => 
		Implementation.S54,
}
\end{lstlisting}
\captionof{lstlisting}{Generated Type for \trole{Svr} Receive State
in \tprotocol{Adder} protocol}
\label{lst:addersvrreceivehandler}
\end{figure}

@@

The interface values are functions parameterised by
the message payload, and must return the successor state encoding.
We see another benefit of defining the payload in Message Types
as a tuple: we can define the receive handler parameter
using the \textit{spread syntax}, which allows the tuple
expression to be expanded into a list of function arguments. 
More concretely, as shown in \cref{lst:nodeefsmspread},
it allows the developer to pattern match on the
individual payload values (\cref{line:yesspread}) 
rather than defining their function to expect a tuple 
and manually destructing it (\cref{line:nospread}),
so the former is more intuitive.

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript,tabsize=2]
const withSpread = (x: number, y: number) => {...} (*@\label{line:yesspread}@*)
const withoutSpread = (payload: [number, number]) => {...} (*@\label{line:nospread}@*)

@@

The behaviour of the runtime is dependent on the current state,
so it needs a way to distinguish between
all the different states -- one can think of this as implementing
the state transition function from the theory, which is analogue to 
overloading a \texttt{next()} method for each state.
Due to limitations in the TypeScript language, this would have to
be some sort of switch statement, with the \texttt{next()}
method parameterised by some base type assignable to all states.
Currently, the state is only determined by the handler
to be implemented by the developer, so the switch statement
and base type would have to be defined on the handler APIs.

@@

\subsubsection{Distinguishing Handlers using Conditional Types}

We attempt to address this by defining an enum of state identifiers
for each type of state (i.e. an enum for send states, 
an enum for receive states)
upon which to execute the EFSM, which solves the switch statement
problem.
Now, we are left with defining a mapping between the 
state identifier enum to the handler type. This construct would be
analogue to \textit{dependent types}, which again, is not a feature
of the TypeScript type system.

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
enum SendState { S1, S3, ... };
enum ReceiveState { S2, S4, ... };
type State = SendState | ReceiveState;

type SendHandler<S extends SendState> = 
	S extends SendState.S1 ? Handler.S1 :
	S extends SendState.S3 ? Handler.S3 : ... ;

type ReceiveHandler<S extends ReceiveState> = 
	S extends ReceiveState.S2 ? Handler.S2 :
	S extends ReceiveState.S4 ? Handler.S4 : ... ;
\end{lstlisting}
\captionof{lstlisting}{Approximating Type Dependency
using \textit{Conditional Types}}
\label{lst:conditionaltypes}
\end{figure}

@@

Unfortunately, this approach does not work for the simple fact that
conditional types were not designed to be exploited in this manner.
The main limitation of conditional types is its 
\textit{distributivity} when the type ``parameter'' is an
union type (which is the case for enums, as
\texttt{S = SendState.S1 | SendState.S3 | ...}), where

@@

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
abstract class ISend { type: 'Send' = 'Send'; ... }
abstract class IReceive { type: 'Receive' = 'Receive'; ... }
abstract class ITerminal { type: 'Terminal' = 'Terminal'; ... }

@@

	export class S51 extends IReceive {
		constructor(private handler: Handler.S51) { super(); }
		...
	}
	...

};
\end{lstlisting}
\captionof{lstlisting}{Discriminated Unions in EFSM 
for Server-Side Endpoints}
\label{lst:nodeefsmimplementation}
\end{figure}

>>>>

We show the structure of the generated \filename{EFSM.ts}
file in
\cref{lst:nodeefsmfile}.
Note that the formal definition of the EFSM in 
\cref{subsection:bgscribble}
contains more than just states and the state transition function,
so we encode the additional information as well.
Each type of information is grouped into their own
\textit{namespace}, and are collectively exported in
the EFSM \textit{module} for the developer to use.

\begin{figure}
\begin{lstlisting}[language=javascript,tabsize=2,title=EFSM.ts]
// (*@\cref{subsection:nodeefsmroleslabelsmsg}@*)
export namespace Roles {...};			// for roles in protocol
export namespace Labels {...};		// for message labels
export namespace Message {...};		// for message structure

// (*@\cref{subsection:nodeefsmhandlers}@*)
export namespace Handler {...};		// for handler APIs

// (*@\cref{subsection:nodeefsmimplementation}@*)
abstract class ISend {...};
abstract class IReceive {...};
abstract class ITerminal {...};
export namespace Implementation {...};
\end{lstlisting}
\captionof{lstlisting}{Structure of Generated EFSM Encoding 
for Server Endpoint}
\label{lst:nodeefsmfile}
\end{figure}

@@

\subparagraph{Messages}
The handler APIs that we generate for developers
need to refer to the label identifier and payload type: 
we refer to this as the message structure.
Each message structure is expressed as an interface with
properties for the label and payload.
These interfaces are grouped based on the EFSM state
they belong using \textit{union types}.
We illustrate this in \cref{lst:addersvrmsg}.
By expressing the payload type as a \textit{tuple}\footnote{
In TypeScript, a tuple is an array with fixed size
and known types for elements at each position.
},
we easily generalise our type definition to polyadic payloads.

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript]
// Inside the Message namespace...
export interface S54THANKS {
	label: Labels.S54.THANKS,
	payload: [string],
};
export interface S54TERMINATE {
	label: Labels.S54.TERMINATE,
	payload: [],
};

@@

Handler APIs act as the \textit{seams} of the EFSM.
A seam is ``a place where one can alter the
behaviour in the program without editing in that place''
\cite{seam}. The developer implements handlers
to define what message to send and how to process a received
message, without access to the send and receive APIs
that are used to execute the EFSM.
We collect these APIs under the \texttt{Handler} namespace. 
We introduce the generated handlers for sending and receiving states.
These are non-terminal states that will involve the encoding of its
successor state. The reader will notice that, in the listings below,
the successor state is stated to be under the
\texttt{Implementation} namespace: we explain in 
\cref{subsection:nodeefsmimplementation}, but for now,
it is sufficient to acknowledge that those refer to the encoding
of the successor state.
As a design choice, we \textit{do not} generate handlers for
terminal states, because the semantics of inactivity mean
there is nothing to handle.

\subparagraph{Send}
We model selections using a union type to
encapsulate the possible send actions, as shown in 
\cref{lst:addersvrsendhandler}.
Each send action is encoded as a tuple of
the label, the payload, and the successor state encoding.
We see some benefits from defining Message Types as interfaces:
TypeScript supports \textit{index type queries} to extract
named property types, so
\lstonelinejs{Message.S54THANKS['payload']} 
would resolve to \lstonelinejs{[string]},
based on the interface definition in \cref{lst:addersvrmsg}.

@@

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript,tabsize=2]
// Inside the Handler namespace...
export interface S51 {
	[Labels.S51.ADD]: (...payload: Message.S51ADD['payload']) =>
		Implementation.S53,
	[Labels.S51.QUIT]: (...payload: Message.S51QUIT['payload']) => 
		Implementation.S54,
}
\end{lstlisting}
\captionof{lstlisting}{Generated Type for \trole{Svr} Receive State
in \tprotocol{Adder} protocol}
\label{lst:addersvrreceivehandler}
\end{figure}

@@

The interface values are functions parameterised by
the message payload, and must return the successor state encoding.
We see another benefit of defining the payload 
in the message structure interface
as a tuple: we can define the receive handler parameter
using the \textit{spread syntax}, which allows the tuple
expression to be expanded into a list of function arguments. 
More concretely, as shown in \cref{lst:nodeefsmspread},
it allows the developer to pattern match on the
individual payload values (\cref{line:yesspread}) 
rather than defining their function to expect a tuple 
and manually destructing it (\cref{line:nospread}),
so the former is more intuitive.

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript,tabsize=2]
// More intuitive
const withSpread = (x: number, y: number) => {...} (*@\label{line:yesspread}@*)

// Needs manual destructuring
const withoutSpread = (payload: [number, number]) => {...} (*@\label{line:nospread}@*)

@@

The behaviour of the runtime is dependent on the current state,
so the runtime needs a way to distinguish between
all the different states -- one can think of this as implementing
the state transition function from the theory, which is analogue to 
overloading a \texttt{next()} method for each state.
TypeScript does not support method overloading,
so the \texttt{next()} method would need to be defined
on some base type assignable to all states,
and the implementation of the \texttt{next()} method 
can use a switch statement to distinguish between the different states.
Currently, the state is only determined by the handler
to be implemented by the developer, so the switch statement
and base type would have to be defined on the handler APIs.

@@

\subsubsection{Distinguishing Handlers using Conditional Types}
\label{subsubsection:dependenttypes}

We attempt to address this by defining an enum of state identifiers
for each type of state (i.e. an enum for send states, 
an enum for receive states)
upon which to execute the EFSM. 
This can solve the switch statement problem.
Now, we are left with defining a mapping between the 
state identifier enum to the handler type. This construct would be
analogue to \textit{dependent types}, which again, is not a feature
of the TypeScript type system.

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
enum SendState { S53, S54 };
enum ReceiveState { S51 };
type State = SendState | ReceiveState;

type SendHandler<S extends SendState> = 
	S extends SendState.S53 ? Handler.S53 :
	S extends SendState.S54 ? Handler.S54 : never ;

type ReceiveHandler<S extends ReceiveState> = 
	S extends ReceiveState.S51 ? Handler.S51 : never;
\end{lstlisting}
\captionof{lstlisting}{Approximating Type Dependency
using \textit{Conditional Types}}
\label{lst:conditionaltypes}
\end{figure}

@@

Unfortunately, this approach does not work for the simple fact that
conditional types were not designed to be exploited in this manner.
The main limitation of conditional types is its 
\textit{distributivity} when the type ``parameter'' is an
union type (which is the case for enums, as
\texttt{S = SendState.S53 | SendState.S54| ...}), where

@@

We illustrate this with an example below.
A \texttt{Shape} is either a \texttt{Circle} or \texttt{Square}.
Both define a property name \texttt{type}, which is the discriminant
property, along with properties unique to their shape
(i.e. radius for circles, side lengths for squares).
The \texttt{area()} function uses the discriminant property
(here, in a switch statement) to narrow the type of \texttt{s}
and perform operations specific to that shape.

\begin{lstlisting}[language=javascript]
// The `type` property is the discriminant
interface Circle { type: 'Circle', radius: number };
interface Square { type: 'Square', side: number };
type Shape = Circle | Square;

function area(s: Shape) {
	switch (s.type) {
		case 'Circle': return 3.14 * s.radius;	// s: Circle
		case 'Square': return s.side * s.side;	// s: Square
	}
}
\end{lstlisting}

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
// The `type` property is the discriminant
abstract class ISend { type: 'Send' = 'Send'; ... } (*@\label{line:nodeefsmsend}@*)
abstract class IReceive { type: 'Receive' = 'Receive'; ... } (*@\label{line:nodeefsmreceive}@*)
abstract class ITerminal { type: 'Terminal' = 'Terminal'; ... } (*@\label{line:nodeefsmterminal}@*)

@@

	export class S51 extends IReceive {

		// Stores the handler defined by the developer
		// as a private property	
		constructor(private handler: Handler.S51) { super(); }
		...
	}
	...	
};
\end{lstlisting}
\captionof{lstlisting}{Discriminated Unions in EFSM 
for Server-Side Endpoints}
\label{lst:nodeefsmimplementation}
\end{figure}

### 2020-06-17 report/sections/server/node/runtime.tex
Report: fix subfig numbering
Report: review chapter 4

We define the session runtime for the \trole{Svr} endpoint
of the \tprotocol{Adder} protocol in Svr.ts,
named after the endpoint.
It exposes a \textbf{public API}
with seams for the developer to pass in the WebSocket server 
and application logic
(i.e. their handler implementations).
It is developer's responsibility to construct the
WebSocket server and set it up to listen for incoming connections.
Internally, it keeps a \textit{private API}
for executing the EFSM, when all participants have 
joined the session.

\begin{lstlisting}[language=javascript]
export class Svr {
	constructor(wss: WebSocket.Server,
				initialState: Implementation.S51) { ... }
	...
}

@@

const roleToSocket: Partial<RoleToSocket> = { (*@\label{line:initroletosocket}@*)
	[Roles.Peers.Client]: undefined,
};

const onSubscribe = ({ data, target: socket }) => {
	const { connect: role } = JSON.parse(data) 
		as Message.ConnectRequest; (*@\label{line:connectionrequest}@*)
	if (!waiting.has(role)) { return socket.close(); } (*@\label{line:occupied}@*)

	roleToSocket[role] = socket; (*@\label{line:connectwsbind}@*)
	waiting.delete(role);

	if (waiting.size === 0)
		new Session(wss, roleToSocket as RoleToSocket, initialState); (*@\label{line:newsession}@*)
};

wss.addEventListener('connection', ws => {
	ws.onmessage = onSubscribe; (*@\label{line:wslisteneroverride}@*)
});
\end{lstlisting}
\captionof{lstlisting}{Handling Connections in Server Endpoint}
\label{lst:nodeconnect}
\end{figure}

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
next(state: Implementation.Type) {
	switch (state.type) {
		case 'Send': {
			const [label, payload, succ]: (*@\hl{???}@*) = state.handler;	 (*@\label{line:noderuntimebadtype1}@*)
			this.send((*@\hl{???}@*), label, payload); (*@\label{line:noderuntimeunknownrole}@*)		
			return this.next(succ);
		}
		case 'Receive': {
			this.wss.onmessage = ({ data }) => {
				const { label, payload } = JSON.parse(data) as (*@\hl{???}@*); (*@\label{line:noderuntimebadtype2}@*)
				const succ: (*@\hl{???}@*) = state.handler[label](...payload); (*@\label{line:noderuntimebadtype3}@*)
				return this.next(succ);
			}
		}
		case 'Terminate': { return; }
	}
}
\end{lstlisting}
\captionof{lstlisting}{Conceptual EFSM Transition Function 
for Server-Side Endpoint}
\label{lst:noderuntimesimple}
\end{figure}

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
next(state: Implementation.Type) {
	switch (state.type) {
		case 'Send':
			return state.performSend(this.next, this.send);
		case 'Receive':
			return state.prepareReceive(this.next, this.register);
		case 'Terminal':
			return;
	}
}
\end{lstlisting}
\captionof{lstlisting}{Final EFSM Transition Function for Server-Side
Endpoint}
\label{lst:noderuntime}
\end{figure}

@@

We get the label, payload and successor implementation
directly from the handler implemented by the developer, 
accurately typed by how we define the handler API in EFSM.ts.
The developer does not need to specify which role
to send to: this is a sensible design choice, as we know this
from the Scribble protocol, so we do not need the developer 
to specify separately.
As a result, we generate the code to send the message
to the correct role (\cref{line:nodesendrole}).
We use the \texttt{send()} method passed down
by the runtime to commit our communication action:
the runtime will handle how to serialise the message and perform
the send. We guarantee that \texttt{send()} is called
\textbf{exactly once} by construction, thus channel linearity
is never violated.
Finally, we use the parameterised EFSM transition handler
to notify the rutime which specific state to transition to.

@@

class Session {
	private handler?: MessageHandler; (*@\label{line:noderuntimehandler}@*)

	constructor(...) {
		...
		Object.values(this.roleToSocket).
			.forEach(ws => ws.onmessage = this.receive.bind(this)); (*@\label{line:wsbind}@*)
		this.handler = undefined;
	}

	register(handler: MessageHandler) { this.handler = handler; } (*@\label{line:register}@*)

	receive({ data }: WebSocketMessage) {
		const handler = this.handler; (*@\label{line:callhandlerstart}@*)
		this.handler = undefined; (*@\label{line:callhandler2}@*)
		handler?.(data); (*@\label{line:callhandler}@*)
	}
}
\end{lstlisting}
\captionof{lstlisting}{Approach to Dynamic WebSocket Message Event Listener}
\label{lst:noderuntimewsmsg}
\end{figure}

@@

	prepareReceive(
		next: EfsmTransitionHandler,
		register: (from: Roles.Peers, messageHandler: MessageHandler)
			=> void
	) {
		const messageHandler = (message: any) => {
			const decoded = JSON.parse(message) as Message.S51;
			switch (decoded.label) {
				case Labels.S51.ADD: {
					const successor = 
						this.handler[decoded.label](...decoded.payload);
					return next(successor);
				}
				case Labels.S51.QUIT: {
					const successor = 
						this.handler[decoded.label](...decoded.payload);
					return next(successor);	
				}
			}            
		}
		register(Roles.Peers.Client, messageHandler);
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Code for \texttt{Implementation} API for
Receive State}
\label{lst:nodereceive}
\end{figure}

@@

\begin{figure}[!h]
\centering
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive2}
\caption{Message processed before transitioning to receive state}
\label{subfig:nodereceivemsgfirst}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive1}
\caption{Message processed after transitioning to receive state}
\label{subfig:nodereceivehandlefirst}
\end{subfigure}
\captionof{figure}{Possible Orderings for Handling Message Event and Preparing
Receive State}
\label{fig:nodereceivecompare}
\end{figure}

@@

\begin{enumerate}
\item We construct types for these two mappings 
(\cref{line:mapped1,,line:mapped2}), 
using a generic mapped typed defined in
\cref{lst:noderolesmappedtypes}.

@@

		this.messageQueue = { [Roles.Peers.Client]: [], }; (*@\label{line:queue1}@*)
		this.handlerQueue = { [Roles.Peers.Client]: [], }; (*@\label{line:queue2}@*)

		this.next(initialState);	
	}

	receive(from: Roles) { (*@\label{line:receivefunc}@*)
		return ({ data }) => { (*@\label{line:receivefuncret}@*)
			// Array.shift() can return undefined if empty
			const handler = this.handlerQueue[from].shift();
			if (handler !== undefined) {
				handler(data);
			} else {
				this.messageQueue[from].push(data);			
			}
		}
	}

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript]
// Inside the Roles namespace...
export type PeersToMapped<Value> = { [Role in Peers]: Value };
\end{lstlisting}
\captionof{lstlisting}{Generic Mapped Type for Roles}
\label{lst:noderolesmappedtypes}
\end{figure}

>>>>

We define the session runtime for the \trole{Svr} endpoint
of the \tprotocol{Adder} protocol in \filename{Svr.t}s,
named after the endpoint.
It exposes a \textbf{public API}
with seams for the developer to pass in the WebSocket server 
and application logic
(i.e. their handler implementations).
It is developer's responsibility to construct the
WebSocket server and set it up to listen for incoming connections.
Internally, it keeps a \textit{private API}
for executing the EFSM, when all participants have 
joined the session.

\begin{lstlisting}[language=javascript]
// Exported to developer
export class Svr {
	constructor(wss: WebSocket.Server,
				initialState: Implementation.S51) { ... }
	...
}

@@

// Mapping of roles to WebSocket connections
const roleToSocket: Partial<RoleToSocket> = { (*@\label{line:initroletosocket}@*)
	[Roles.Peers.Client]: undefined,
};

// Invoked when a connection request is received
const onSubscribe = ({ data, target: socket }) => {

	// Deserialise connection request message
	const { connect: role } = JSON.parse(data) 
		as Message.ConnectRequest; (*@\label{line:connectionrequest}@*)

	// Ignore if role is aleady taken
	if (!waiting.has(role)) { return socket.close(); } (*@\label{line:occupied}@*)

	// Map the role in the connection request to this WebSocket
	roleToSocket[role] = socket; (*@\label{line:connectwsbind}@*)
	waiting.delete(role);

	// Start executing EFSM when all roles have joined
	if (waiting.size === 0) {
		new Session(wss, roleToSocket as RoleToSocket, initialState); (*@\label{line:newsession}@*)
	}
};

// For every new connection, process message with `onSubscribe`
wss.addEventListener('connection', ws => {
	ws.onmessage = onSubscribe; (*@\label{line:wslisteneroverride}@*)
});
\end{lstlisting}
\captionof{lstlisting}{Handling Connections in Server Endpoint}
\label{lst:nodeconnect}
\end{figure}

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
next(state: Implementation.Type) {
	// Distinguish between states using discriminant property
	switch (state.type) {
		case 'Send': {
			const [label, payload, succ]: (*@\hl{???}@*) = state.handler;	 (*@\label{line:noderuntimebadtype1}@*)
			this.send((*@\hl{???}@*), label, payload);	// Who to send to? (*@\label{line:noderuntimeunknownrole}@*)		
			return this.next(succ);
		}
		case 'Receive': {
			// Handle incoming messages using this anonymous function
			this.wss.onmessage = ({ data }) => {

				// Which message structure to use to deserialise?
				const { label, payload } = JSON.parse(data) as (*@\hl{???}@*); (*@\label{line:noderuntimebadtype2}@*)
				const succ: (*@\hl{???}@*) = state.handler[label](...payload); (*@\label{line:noderuntimebadtype3}@*)
				return this.next(succ);
			}
		}
		case 'Terminate': { return; }
	}
}
\end{lstlisting}
\captionof{lstlisting}{Conceptual EFSM Transition Function 
for Server-Side Endpoint}
\label{lst:noderuntimesimple}
\end{figure}

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
next(state: Implementation.Type) {
	switch (state.type) {
		case 'Send':				// recall (*@\cref{lst:nodeefsmimplementation} \cref{line:nodeefsmsend}@*)
			return state.performSend(this.next, this.send);
		case 'Receive':			// recall (*@\cref{lst:nodeefsmimplementation} \cref{line:nodeefsmreceive}@*)
			return state.prepareReceive(this.next, this.register);
		case 'Terminal':		// recall (*@\cref{lst:nodeefsmimplementation} \cref{line:nodeefsmterminal}@*)
			return;
	}
}
\end{lstlisting}
\captionof{lstlisting}{Final EFSM Transition Function for Server-Side
Endpoint}
\label{lst:noderuntime}
\end{figure}

@@

We get the label, payload and successor implementation
directly from the handler implemented by the developer, 
accurately typed by how we define the handler API in \filename{EFSM.ts}.
The developer does not need to specify which role
to send to: this is a sensible design choice, as we know this
from the Scribble protocol, so we do not need the developer 
to specify separately.
As a result, we generate the code to send the message
to the correct role (\cref{line:nodesendrole}).
We use the \texttt{send()} method passed down
by the runtime to commit our communication action:
the runtime will handle how to serialise the message and perform
the send. We guarantee that \texttt{send()} is called
\textbf{exactly once} by construction, thus channel linearity
is never violated.
Finally, we use the parameterised EFSM transition handler
to notify the rutime which specific state to transition to.

@@

class Session {
	// Optional type, same as `MessageHandler | undefined`
	private handler?: MessageHandler; | undefined(*@\label{line:noderuntimehandler}@*)

	constructor(...) {
		...
		// Process incoming messages from each WebSocket
		// using `this.receive()`
		Object.values(this.roleToSocket).
			.forEach(ws => ws.onmessage = this.receive.bind(this)); (*@\label{line:wsbind}@*)

		// Initialise handler as undefined
		this.handler = undefined;
	}

	// Set the handler to be used to handle the next receive event
	register(handler: MessageHandler) { this.handler = handler; } (*@\label{line:register}@*)

	receive({ data }: WebSocketMessage) {
		const handler = this.handler; (*@\label{line:callhandlerstart}@*)

		// `Unregister` the current receive handler
		this.handler = undefined; (*@\label{line:callhandler2}@*)

		// `handler` has an optional type, so we first
		// check if there is a value set, then invoke the handler.
		// Same as `if (handler !== undefined) { handler(data); }`
		handler?.(data); (*@\label{line:callhandler}@*)
	}
}
\end{lstlisting}
\captionof{lstlisting}{Attempt to Dynamic WebSocket Message Event Listener}
\label{lst:noderuntimewsmsg}
\end{figure}

@@

	prepareReceive(
		next: EfsmTransitionHandler,
		register: (from: Roles.Peers, messageHandler: MessageHandler)
			=> void
	) {
		// Define dynamic WebSocket message event listener
		const messageHandler = (message: any) => {

			// Deserialise message
			const decoded = JSON.parse(message) as Message.S51;

			// Discriminate between messages using label
			switch (decoded.label) {
				case Labels.S51.ADD: {
					// Invoke handler to get successor state
					const successor = 
						this.handler[decoded.label](...decoded.payload);

					// Invoke callback to advance EFSM
					return next(successor);
				}
				case Labels.S51.QUIT: {
					const successor = 
						this.handler[decoded.label](...decoded.payload);
					return next(successor);	
				}
			}            
		}

		// Register the message handler under the
		// WebSocket bound to the `Client` role
		register(Roles.Peers.Client, messageHandler);
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Code for \texttt{Implementation} API for
Receive State}
\label{lst:nodereceive}
\end{figure}

@@

\begin{figure}[!h]
\centering
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive2}
\caption{Message processed before transitioning to receive state}
\label{subfig:nodereceivemsgfirst}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive1}
\caption{Message processed after transitioning to receive state}
\label{subfig:nodereceivehandlefirst}
\end{subfigure}
\caption{Possible Orderings for Handling Message Event and Preparing
Receive State}
\label{fig:nodereceivecompare}
\end{figure}

@@

\begin{enumerate}
\item We construct types for these two mappings 
(\cref{line:mapped1,,line:mapped2}), 
using a generic mapped typed defined below.
\begin{lstlisting}[language=javascript]
// Inside the Roles namespace...
export type PeersToMapped<Value> = { [Role in Peers]: Value };
\end{lstlisting}

@@

		// Set up empty queues
		this.messageQueue = { [Roles.Peers.Client]: [], }; (*@\label{line:queue1}@*)
		this.handlerQueue = { [Roles.Peers.Client]: [], }; (*@\label{line:queue2}@*)

		this.next(initialState);		// Advance to initial state
	}

	receive(from: Roles) { (*@\label{line:receivefunc}@*)

		// Return a WebSocket message event listener
		// that looks at the message/handler queue
		// specific to the `from` role

		return ({ data }) => { (*@\label{line:receivefuncret}@*)
			// Array.shift() can return undefined if empty
			const handler = this.handlerQueue[from].shift();
			if (handler !== undefined) {
				handler(data);
			} else {
				this.messageQueue[from].push(data);			
			}
		}
	}

@@
